---
title: "College_Basketball_Data_Analysis"
format: 
  html:
    code-tools: true
    toc: true
    embed-resources: true
editor: source
---


```{r}
#| include: false

library(tidyverse)
library(dagitty)
library(ggplot2)
library(bayesplot)
library(CalvinBayes)
```

```{r}
#| include: false
# load data_set
data <- read_csv("summary25_pt.csv")
```

```{r}
# glimpse of the data
glimpse(data)
```
## Data Source


This data includes metrics such as:

- Tempo (pace of play)

- Offensive/Defensive Efficiency (OE/DE)

- Adjusted metrics (AdjOE, AdjDE)

- Efficiency Margins (AdjEM)

- Seed lines

## Research question

Our goal is to answer the question:

What is the probability that a team ends up in a final four based on their adjusted offensive, defensive efficiency, adjusted efficiency margin, tempo, and seed line?

This will help us to understand which team characteristics are most influential in having to win games in the NCAA tournament and reaching a final four


## Variables of Interest

- Response variable: final_four 

This is a binary variable we will create, indicating whether a team has made a final four (TRUE if final_four = 1, otherwise FASLE)

- Key predictor: AjdEM (AdjOE - AdjDE)

- Other predictors may include AdjOE, AdjDE, AdjTempo, Seed

```{r}
# binary response variable

data <- data %>%
  mutate (final_four = ifelse(final_four <0, FALSE, TRUE))

# check the class balance
table(data$final_four)
```


```{r}
# causal diagram

causal <- dagitty("dag{
  AdjOE -> final_four;
  AdjDE -> final_four;
  AdjOE -> AdjEM;
  AdjDE -> AdjEM;
  AdjTempo -> final_four;
  AdjEM -> final_four;
  Tempo -> AdjTempo;
  OE -> AdjOE;
  DE -> AdjDE;
  AdjOE -> seed;
  AdjDE -> seed;
  seed -> final_four
}")
# plot it
gg_dag(causal, size = 15, highlight = 'final_four' 
#      + theme( plot.margin = margin(15, 15, 15, 15)
  )
```

In the next steps of the analysis, we will:
	
- Fit the model with more iterations to refine the estimates
- Check model diagnostics (convergence check and residuals)
- Fitting the model using MCMC.
- Fit a logistic regression (GLM with binomial distribution) to model 
final_four ~ AdjOE + AdjDE AdjEM + AdjTempo + Seed.
- Interpret which performance metrics most significantly affect a team's chances of making a final four
- Evaluate prediction performance.




