---
title: "College_Basketball_Data_Analysis"
authors: 
  - Blessing Amoah
  - Alex Barnard
  - Fengyi Li
format: 
  html:
    code-tools: true
    toc: true
    embed-resources: true
editor: source
---


```{r}
#| include: false

library(tidyverse)
library(dagitty)
library(ggdag)
library(ggplot2)
library(bayesplot)
library(rethinking)
```

```{r}
#| include: false
# load data_set
data <- read_csv("summary25_pt.csv")
```

```{r}
# glimpse of the data
glimpse(data)
```
## Data Source


This data includes metrics such as:

- Tempo (pace of play)

- Offensive/Defensive Efficiency (OE/DE)

- Adjusted metrics (AdjOE, AdjDE)

- Efficiency margins

- Team seeds which indicates their tournament ranking.

## Research question

Our goal is to answer the question:

What is the probability that a team ends up in the Final Four based on their adjusted offensive and defensive efficiency?

This will help us understand which team characteristics are most influential in achieving Final Four success, which is critical for predicting tournament outcomes.


## Variables of Interest

- Response variable: Final_four

This is a binary variable indicating whether a team made it to the Final Four (TRUE) or not (FALSE)

- Key predictor: AdjOE thus Adjusted Offensive Efficiency

- Other predictors may include AdjDE, AdjTempo, AdjEM(Efficiency Margin).

```{r}
# Convert Final Four appearance to binary (TRUE if team reached Final Four, FALSE otherwise)

data <- data %>%
  mutate(final_four = ifelse(final_four >= 0, TRUE, FALSE))

# check the class balance
table(data$final_four)
```


```{r}
# causal diagram

offensive_defensive_metrics <- dagitty("dag{
  AdjOE -> Final_Four;
  AdjDE -> Final_Four;
  AdjOE -> AdjEM;
  AdjDE -> AdjEM;
  AdjTempo -> Final_Four;
  AdjEM -> Final_Four;
  Tempo -> AdjTempo;
  OE -> AdjOE;
  DE -> AdjDE;
  AdjOE -> seed;
  AdjDE -> seed;
  seed -> Final_Four
}")
# plot it
ggdag(offensive_defensive_metrics) +
  theme(
    plot.margin = margin(15, 15, 15, 15)
  )
```


In the next steps of the analysis, we will:
	
- Fit the model with more iterations to refine the estimates
- Check model diagnostics (convergence check and residuals)
- Fitting the model using MCMC.
- Fit a logistic regression (GLM with binomial distribution) to model seed_top4 ~ AdjOE + AdjDE + AdjTempo + AdjEM.
- Interpret which performance metrics most significantly affect a team’s seeding.
- Evaluate prediction performance.



