---
title: "Prior Predictive Dist"
format: html
---

```{r}
#| include: false
library(dplyr)    
library(tidyverse)
library(ggformula)
library(rstan)
library(rethinking)
library(CalvinBayes)
library(tidybayes)
library(bayesplot)
```


```{r}
#| include: false
# load data_set
dat <- read_csv("all_seasons.csv")
```

```{r}
data <- dat %>% mutate(final_four = ifelse(is.na(final_four), 0, final_four))
```


```{r}
# Convert Final Four appearance to binary (TRUE if team reached Final Four, FALSE otherwise)

data <- data %>%
  mutate(final_four = ifelse(final_four >= 0, TRUE, FALSE))


# check the class balance
table(dat$final_four)
```

```{r}
clean_data <- data %>%
  drop_na(seed, AdjOE, AdjDE, AdjEM, AdjTempo)
nrow(clean_data)

```

```{r}
dat <- list(
  final_four = as.integer(clean_data$final_four),
  AdjOE = scale(clean_data$AdjOE),
  AdjDE = scale(clean_data$AdjDE),
 AdjEM = scale(clean_data$AdjEM),
  AdjTempo = scale(clean_data$AdjTempo),
 seed = clean_data$seed
 
#  N = nrow(clean_data),
#  N_seeds = max(clean_data$seed, na.rm = TRUE)
)

```


```{r}
n_sim <- 10
sigma_seed = dexp(1)

prior_pred_dists <- tibble(
  b1 = rnorm(n_sim, mean = 0, sd = 1.5),
  b2 = rnorm(n_sim, mean = 0, sd = 1.5),
  b3 = rnorm(n_sim, mean = 0, sd = 1.5),
  b4 = rnorm(n_sim, mean = 0, sd = 1.5),
  b_seed = rnorm(n_sim, mean = 0, sigma_seed),
  sim_number = c(1:n_sim)
  )

# what did we do so far?
# to make sure you understand, have a peek at the current result...
glimpse(prior_pred_dists)

```



```{r}
#| include: false

target_probability = (4/68)

m_mcmc <- ulam(
  alist(
    final_four ~ dbinom(1, p),
    logit(p) <- b1*AdjOE + b2*AdjDE + b4*AdjTempo + b5*seed,
    


    # Priors for main effects
    b1 ~ dnorm(logit(.1), .2), #AdjOE
    b2 ~ dnorm(logit(.15), .2), #AdjDE
    b5 ~ dnorm(0, .2), #seed
    b4 ~ dnorm(-logit(.2), .2) #temp

 
  ),
  data = dat,
  chains = 4,
  cores = 4,
  iter = 2000
)
```

```{r}
precis(m_mcmc, depth =  2)
traceplot(m_mcmc)
```



```{r}
final_four_post_samp <- as.data.frame(m_mcmc)
```






```{r}
set.seed = 1
# Set number of simulations
n_sim <- 1000 # 
target_probability = (4/68) #only 200 out of the 3400 teams will make a final four

# Simulate prior values for coefficients
b1 <- rnorm(n_sim, logit(.1), .2) # 1 sd in AjdOE gives a 10% increase of making a final four
b2 <- rnorm(n_sim, logit(.15), .2)
b4 <- rnorm(n_sim, 0, .2)
b5 <- rnorm(n_sim, -logit(.2), .2)

# Simulate group-level effects

# Simulate standardized predictor values
AdjOE <- rnorm(n_sim)
AdjDE <- rnorm(n_sim)
AdjTempo <- rnorm(n_sim)
b_seed <- rnorm(n_sim)

# Linear predictor
logit_p <- b1 * AdjOE + b2 * AdjDE + b4 * AdjTempo + b5 * b_seed

logit_target <- logit(target_probability) 

logit_p <- logit_p - (mean(logit_p) - logit_target) #scale down logit to more accurately reflect only 200 teams making it

# Transform to probability
p <- 1 / (1 + exp(-logit_p))


# Simulate binary final_four outcome
final_four_sim <- rbinom(n_sim, size = 1, prob = p)

# Make dataframe for plotting
prior_sim_df <- tibble(
  probability = p,
  outcome = final_four_sim
)

# Plot histogram of prior predictive probabilities
ggplot(prior_sim_df, aes(x = probability)) +
  geom_histogram(bins = 100, fill = "maroon", color = "gold") +
  labs(title = "Prior Predictive Distribution for Final Four",
       x = "Simulated Probability (Prior Only)",
       y = "Count") +
  theme_minimal()
```



```{r}
# Teams to simulate per prior draw
n_teams <- 100  

# prior_pred_dists for each team
prior_pred_df <- prior_pred_dists %>%
  slice(rep(1:n(), each = n_teams)) %>%
  mutate(
    team_id = rep(1:n_teams, times = nrow(prior_pred_dists)),
    AdjOE = rnorm(n(), mean = 0, sd = 1),
    AdjDE = rnorm(n(), mean = 0, sd = 1),
    AdjEM = rnorm(n(), mean = 0, sd = 1),
    AdjTempo = rnorm(n(), mean = 0, sd = 1)
  ) %>%
  mutate(
    logit_p = b1 * AdjOE + b2 * AdjDE + b4 * AdjTempo + b5 * b_seed,
    logit_p <- logit_p - (mean(logit_p) - logit_target),
    p = 1 / (1 + exp(-logit_p)),
    final_four_sim = rbinom(n(), size = 1, prob = p)
  )
```


```{r}
# Prior predictive probability
gf_point(p ~ team_id | sim_number, 
         data = prior_pred_df,
         alpha = 0.5) +
  labs(
    title = "Prior Predictive Simulations (Final Four Probability)",
    x = "Team ID",
    y = "Simulated Probability (Prior Only)"
  )
```


### Interpretation of the Prior Predictive Distribution

Without utilizing any observed data, the prior predictive distribution shows the likelihood that a team would advance to the Final Four based only on samples from the previous distributions of the model parameters.

#### Key Takeaways:

- The histogram demonstrates that, under the prior alone, the odds of making it to the Final Four fluctuate throughout the whole [0, 1] range, with a notable bulk near **0 and 1**.

- The faceted scatterplot (one panel per simulation) illustrates how the probability of the Final Four is impacted by preconceived notions about team metrics (such as Adjusted Offensive and Defensive Efficiency) for several hypothetical teams.

- The distribution's shape indicates that the priors used (`Exponential(1)` for seed-level variation and `Normal(0,1.5)` for coefficients) are **weakly informative** but **flexible enough** to permit both high and low probability based on the simulated covariates.

- The probability spread demonstrates that the model is not overconfident before to data viewing and that, when updated via posterior estimation, the model can learn from the data.

Because it confirms that our priors encode appropriate uncertainty and do not compel the model to make irrational assumptions, this phase is crucial to Bayesian modeling.  Overly narrow priors may have been indicated if the majority of probability had clustered firmly around 0.5.  They would indicate overconfident, unrealistic priors if they were primarily at 0 or 1.

#### Conclusion:
Our prior predictive check confirms that the selected priors are suitable for learning from the real data in the following stage of the study since they achieve a decent balance between **regularization** and **flexibility**.
